### 2018-03-15 paper/3-abstractions.tex
Revise as discussed

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad (from Control.Monad.State)
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
\end{minted}
\begin{minted}[fontsize=\small]{haskell}
newtype Const ks a = Const { getConst :: ks }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const ks) where
    fmap _ (Const ks) = Const ks
instance Monoid ks => Applicative (Const ks) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-2mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems.
% that allows
% us to abstract away from specific application domains, such as software
% development or spreadsheets.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up-to-date. The store is a key-value map, together with \emph{persistent build
information} that is maintained by the build system itself.

\paragraph{Keys and values.} In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.

\paragraph{Input, output, and intermediate values}
Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

\paragraph{The store.}
A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells).

\paragraph{Hashes.}
We use a cryptographic \emph{hash function}
\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
efficient tracking and sharing of build results, where \hs{Hash} is an abstract
data type equipped with the equality test.
In addition to usual \hs{getValue} and \hs{putValue} operations,
some stores support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}\footnote{For
example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
file contents only on demand.}.

\paragraph{Persistent build information.}
The store also holds a value of type \hs{i},
representing the persistent build information maintained by the build
system~--~its ``memory''.

% Given an \hs{i} and a key-value map \store, we can
% \hs{initialise} a store.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\hs{fetch} function and a key. It pattern-matches on the key to see if it has a
task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}.   We will
explain why in the next subsection (\S\ref{sec-general-build}).

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date. What exactly does ``up-to-date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}. Here is a simple build system:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v -- Drop Hashable? It's confusing here
busy task key store = execState (compute key) store
  where
    compute :: k -> State (Store i k v) v
    compute k = case task compute k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system uses a helper function
\hs{compute} that, given a key, brings the key up to date in the store, and returns its value.
The function \hs{compute} runs in the standard Haskell \hs{State} monad -- see
Figure~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{compute} asks the task description \hs{task} how
to compute \hs{k}.
If \hs{task} returns \hs{Nothing} the key is an input, so \hs{compute} simply reads
the result from the store.
Otherwise \hs{compute} runs the action \hs{act}
returned by the \hs{task} to produce a resulting value~\hs{v}, records
the new key/value mapping in the store, and returns \hs{v}.
Notice that \hs{compute} passes itself to \hs{task} as an argument,
so that the latter can use \hs{compute} to recursively find the
values of \hs{k}'s dependencies.

Given an acyclic task description, the \hs{busy} build system terminates with a
correct result, but it is not a \emph{minimal} build system
(Definition~\ref{def-minimal}). Since \hs{busy} has no memory (\hs{i=()}), it
cannot keep track of keys it has already built, and will therefore busily
recompute the same keys again and again if they have multiple dependents. We
will develop much more efficient build systems in~\S\ref{sec-implementations}.

@@

\noindent
As one can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
\hs{compute}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
formula.

\subsection{The need for polymorphism in \hs{Task}}

Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to \hs{State (Store i k v)},
so that the call-back to \hs{fetch :: k -> f v} can side-effect the \hs{Store},
thereby allowing successive calls to \hs{fetch} to communicate with one another.

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description T, we want to explore \emph{many} build systems that
can build T -- and we will do so in Sections~\ref{sec-build} and \ref{sec-implementations}.
As we shall see, each build system will use a different \hs{f}, so the task description
must not fix \hs{f}.

But nor can the task description work for \emph{any} \hs{f}; most
task descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that
\hs{f} satisfies certain properties, such as \hs{Monad} or
\hs{Applicatove}.  That is why \hs{task} has the ``\hs{c f =>}''
constraint in its type, expressing that \hs{f} can only be instantiated by types
that satisfy the constraints \hs{c}.

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task Applicative}. In \hs{sprsh1} we needed only \hs{Applicative} operations,
  expressing the fact that the dependencies between cells can be determined \emph{statically};
  that is, by looking at the formulae, without ``executing'' them (see \S\ref{sec-deps}).
\item \hs{Task Monad}.  As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula
  may depend on the value of cell C, but \emph{which} cell C depends on
  the value of another cell D.
\item \hs{Task Functor} is somehat degenerate: the task description cannot even use
  the application operator \hs{<*>}, which limits dependencies to a single linear chain.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task Monad} as its argument, an application of
\hs{busy} to a \hs{Task Applicative} will typecheck just fine.  It feels a bit
like sub-typing, but it is actually just ordinary higher-rank polymorphism at work.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task Monad k v}. Consider this spreadsheet
example:

@@

\noindent
Here, \cmd{B1} and \cmd{B2} statically depend on each other, but Excel (which
uses dynamic dependencies) is perfectly happy. We can express this using
our task abstraction as follows\footnote{The spreadsheet example that uses
the \hs{INDIRECT} function can be expressed very similarly: simply replace the
line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.}:

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
then \hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

@@

We can now say what it means for a build system to be \emph{correct}, something
that is seldom stated formally. Our intuition is this: \emph{when the build system
  returns, the specified key, and all its dependencies, should be up to date}.
What does ``up to date'' mean?  It means that if we recompute the value of the
key (using the task description, and the final store), we should get exactly
the same value as we see in the final store.

To express this formally we need an auxiliary
function \hs{recompute}, that computes the value of a key in a given store
\emph{without attempting to update any dependencies}:
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}
\noindent
\todo{SLPJ}{Why monad?}
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad.
The use of \hs{Identity} just fixes the `impedance mismatch'
between the pure function \hs{getValue}~\hs{store}, whose type is \store, and
the \hs{fetch} argument of the \hs{task}, whose type must be
\hs{k}~\hs{->}~\hs{f}~\hs{v} for some \hs{f}. To fix the mismatch, we wrap the
result of the pure function in the \hs{Identity} monad: the function
\hs{Identity}~\hs{.}~\hs{getValue}~\hs{store} has the type
\hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a \hs{task}. The
result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now need to get rid
of the \hs{Identity} wrapper by applying \hs{runIdentity} to the contents of
\hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, result :: Store i k v
result = build task key store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{recompute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} will be provided in the next
    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
    analysis. This requirement guarantees that (i) no inputs were corrupted
    during the build, and (ii)~the \hs{ideal} store has constraints not only
    on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

>>>>

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}
\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: Eq k => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash v
initialise :: i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Hashing
hash :: Hashable a => a -> Hash a
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard State monad from Control.Monad.State
data State s a
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard types from Data.Functor.Identity and Data.Functor.Const
newtype Identity a = Identity { runIdentity :: a }
newtype Const m a  = Const    { getConst    :: m }
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Functor (Const m) where
    fmap _ (Const m) = Const m
\end{minted}
\vspace{0.5mm}
\begin{minted}[fontsize=\small]{haskell}
instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{-3mm}
\caption{Signatures of main data types and library functions.}\label{fig-types}
\vspace{-4mm}
\end{figure}
We begin by establishing a common vocabulary for build systems.
% that allows us to abstract away from specific application domains, such as
% software development or spreadsheets.
Fig.~\ref{fig-types} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

\vspace{1mm}
A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up to date. The store is a key/value map, together with \emph{persistent build
information} that is maintained by the build system itself.

\vspace{-2mm}
\paragraph{Keys and values.} In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.

\vspace{-2mm}
\paragraph{Input, output, and intermediate values}
Some values must be provided by the user as \emph{input}. For example,
\cmd{main.c} can be edited by the user who relies on the build system to
compile it into \cmd{main.o} and subsequently \cmd{main.exe}. End build products,
such as \cmd{main.exe}, are \emph{output} values. All other values (in this case
\cmd{main.o}) are \emph{intermediate}; they are not interesting for the user
but are produced in the process of turning inputs into outputs.

% \vspace{-2mm}
% \paragraph{The store.}
A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells). The store also holds a value of type \hs{i},
representing the persistent build information maintained by the build
system~--~its ``memory''.
% Given an \hs{i} and a key/value map \store, we can
% \hs{initialise} a store.

\vspace{-2mm}
\paragraph{Hashes.}
We use a cryptographic \emph{hash function}
\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash}~\hs{v}
for efficient tracking and sharing of build results, where \hs{Hash}~\hs{v} is
an abstract data type equipped with~the~equality test.
In addition to usual \hs{getValue} and \hs{putValue} operations,
some stores support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}\footnote{For
example, Git Virtual File System~\cite{gvfs} stores file hashes and downloads
file contents only on demand.}.

@@

Our first main abstraction is for \emph{task descriptions}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Task Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> fetch "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _     _    = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given a
\emph{callback} \hs{fetch} and a key. It pattern-matches on the key to see if it
has a task description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating the key is an input. If there is a formula in the cell,
it computes the value of the formula, using \hs{fetch} to find the value of any
keys on which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures are given in Fig.~\ref{fig-types}. We will
explain why in subsection~\S\ref{sec-general-build}.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up to date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\noindent
Here is a simple build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: Eq k => Build Monad () k v
busy task key store = execState (fetch key) store
  where
    fetch :: k -> State (Store i k v) v
    fetch k = case task fetch k of
        Nothing  -> do s <- get; return (getValue s k)
        Just act -> do v <- act; modify (\s -> putValue s k v); return v
\end{minted}
\vspace{1mm}

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-types} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{task} how
to compute \hs{k}. If \hs{task} returns \hs{Nothing} the key is an input, so
\hs{fetch} simply reads the result from the store. Otherwise \hs{fetch} runs
the action \hs{act} returned by the \hs{task} to produce a resulting
value~\hs{v}, records the new key/value mapping in the store, and returns \hs{v}.
Notice that \hs{fetch} passes itself to \hs{task} as an argument, so that the
latter can use \hs{fetch} to recursively find the values of \hs{k}'s dependencies.

Given an acyclic task description, the \hs{busy} build system terminates with a
correct result, but it is not a \emph{minimal} build system
(Definition~\ref{def-minimal}). Since \hs{busy} has no memory
(\hs{i}~\hs{=}~\hs{()}), it cannot keep track of keys it has already built, and
will therefore busily recompute the same keys again and again if they have
multiple dependents. We will develop much more efficient build systems
in~\S\ref{sec-implementations}.

@@

\noindent
As one can see, \hs{busy} built both \cmd{B2} and its dependency \cmd{B1} in the
right order (if it had built \cmd{B2} before building \cmd{B1}, the result would
have been $20 * 2 = 40$ instead of $(10 + 20) * 2 = 60$). As an example showing
that \hs{busy} is not minimal, imagine that the formula in cell \cmd{B2} was
\cmd{B1~+~B1} instead of \cmd{B1~*~2}. This would lead to calling
\hs{fetch}~\hs{"B1"} twice -- once per each occurrence of \cmd{B1} in the
formula.

\subsection{The need for polymorphism in \hs{Task}}\label{sec-why-polymorphism}

Now we can see why the \hs{Task} abstraction is polymorphic in \hs{f}:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
The \hs{busy} build system instantiates \hs{f} to
\hs{State}~\hs{(Store}~\hs{i}~\hs{k}~\hs{v)},
so that \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} can side-effect the
\hs{Store}, thereby allowing successive calls to \hs{fetch} to communicate with
one another.

We really, really want \hs{Task} to be \emph{polymorphic} in \hs{f}.
Given \emph{one} task description \cmd{T}, we want to explore \emph{many} build
systems that can build \cmd{T} -- and we will do so in sections~\S\ref{sec-build}
and~\S\ref{sec-implementations}. As we shall see, each build system will use a
different \hs{f}, so the task description must not fix \hs{f}.

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicatove} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}.

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot even
  use the application operator \hs{<*>}, which limits dependencies to a single
  linear chain.
  % It is interesting to note that if \hs{busy} is given a \emph{functorial task}
  % of type \hs{Task}~\hs{Functor}, which cannot use the application operator
  % \hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
  % rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
  % functions, which cannot branch into multiple subproblems.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
argument, an application of \hs{busy} to a \hs{Task}~\hs{Applicative} will
typecheck just fine.  It feels a bit like sub-typing, but it is actually just
ordinary higher-rank polymorphism at work.

@@

As explained in~\S\ref{sec-background-excel}, some task descriptions have dynamic
dependencies, which are determined by values of intermediate computations. In
our framework, such task descriptions correspond to the type
\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}. Consider this spreadsheet example:

@@

\noindent
Here, \cmd{B1} and \cmd{B2} statically depend on each other, but \Excel (which
uses dynamic dependencies) is perfectly happy. We can express this using
our task abstraction as follows:

% The spreadsheet example that uses
% the \hs{INDIRECT} function can be expressed very similarly: simply replace the
% line containing the \cmd{if} statement with \hs{fetch ("A" ++ show c1)}.

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} \emph{and
\hs{fetch} different keys depending on whether or not \hs{c1}~\hs{==}~\hs{1}}.

@@

We can now say what it means for a build system to be \emph{correct}, something
that is seldom stated formally. Our intuition is this: \emph{when the build
system completes, the specified key, and all its dependencies, should be up to
date}. What does ``up to date'' mean? It means that if we recompute the value of
the key (using the task description, and the final store), we should get exactly
the same value as we see in the final store.

To express this formally we need an auxiliary function \hs{compute}, that
computes the value of a key in a given store \emph{without attempting to update
any dependencies}:
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}

Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the pure
function \hs{getValue}~\hs{store}, whose type is \store, and the \hs{fetch}
argument of the \hs{task}, whose type must be \hs{k}~\hs{->}~\hs{f}~\hs{v} for
some \hs{f}. To fix the mismatch, we wrap the result of the pure function in the
\hs{Identity} monad: the function \hs{Identity}~\hs{.}~\hs{getValue}~\hs{store}
has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
contents of \hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = build task key store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of recomputing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} is provided in the Appendix, which
    is dedicated to the task dependency analysis. This requirement guarantees
    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
    store has constraints not only on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}
